### 2018-03-09 paper/background.tex
Restructure the paper

\Excel's recalculation algorithm~\cite{excel_recalc} on the other hand is
significantly different from \Make. Since it is impossible to construct the full
dependency graph upfront in the presence of dynamic dependencies, \Excel
maintains the \emph{calculation chain}, which is an approximation to the correct
topological order produced by the previous build. During recalculation, \Excel
processes cells in this order, but can \emph{defer recalculation of a cell} by
moving it down the chain when one of its newly discovered dynamic dependencies
is dirty. We discuss this algorithm in more detail in~\S\ref{sec-examples}.

@@

\subsection{\Bazel: a cloud build system}
\label{sec-background-shake}

@@

To support cloud builds, \Bazel maintains a conventional \emph{content-addressable
cache}, which can be used to fetch a previously built file given the hash of its
content, as well as \emph{dependency graphs from all previous builds}, annotated
with observed file hashes\footnote{Here we ignore the issue of limited cloud
storage resources for the sake of simplicity; in practice, old entries are
regularly evicted from the storage, as further discussed
in~\S\ref{sec-sec-cloud-aspects}.}. The latter allows to bypass the execution of
a task, by predicting the hash of its result from the hashes of the dependencies,
and subsequently fetch the result from the cache. A concrete implementation is
provided in~\S\ref{sec-examples}.

@@

We summarise differences between four discussed build systems in
Table~\ref{tab-summary}. The `Metadata' column refers to the \emph{auxiliary
build information} persistently stored between builds by all build systems:
\begin{itemize}
    \item \Make stores file modification times, or rather, it relies on the file
    system to do that. In principle a single dirty bit is sufficient, as will be
    demonstrated in~\S\ref{sec-examples}.
    \item \Excel stores the dirty bit as well as the calculation chain.
    \item \Shake stores the dependency graph discovered in the previous build,
    annotated with file content hashes for efficient checking of file changes.
    \item \Bazel stores all dependency graphs discovered in previous builds
    annotated with file hashes, and the content-addressable cache.
\end{itemize}

>>>>

\Excel's recalculation algorithm~\cite{excel_recalc} on the other hand is
significantly different from \Make. Since it is impossible to construct the full
dependency graph upfront in the presence of dynamic dependencies, \Excel
maintains the \emph{calculation chain}, which is an approximation to the correct
topological order produced by the previous build. During recalculation, \Excel
processes cells in this order, but can \emph{defer recalculation of a cell} by
moving it down the chain when one of its newly discovered dynamic dependencies
is dirty. We discuss this algorithm in more detail in~\S\ref{sec-implementations}.

@@

\subsection{\Bazel: a cloud build system}
\label{sec-background-bazel}

@@

To support cloud builds, \Bazel maintains a conventional \emph{content-addressable
cache}, which can be used to fetch a previously built file given the hash of its
content, as well as \emph{dependency graphs from all previous builds}, annotated
with observed file hashes\footnote{Here we ignore the issue of limited cloud
storage resources for the sake of simplicity; in practice, old entries are
regularly evicted from the storage, as further discussed
in~\S\ref{sec-cloud-aspects}.}. The latter allows to bypass the execution of
a task, by predicting the hash of its result from the hashes of the dependencies,
and subsequently fetch the result from the cache. A concrete implementation is
provided in~\S\ref{sec-implementations}.

@@

We summarise differences between four discussed build systems in
Table~\ref{tab-summary}. The `Metadata' column refers to the \emph{auxiliary
build information} persistently stored between builds by all build systems:
\begin{itemize}
    \item \Make stores file modification times, or rather, it relies on the file
    system to do that. In principle a single dirty bit is sufficient, as will be
    demonstrated in~\S\ref{sec-implementations}.
    \item \Excel stores the dirty bit as well as the calculation chain.
    \item \Shake stores the dependency graph discovered in the previous build,
    annotated with file content hashes for efficient checking of file changes.
    \item \Bazel stores all dependency graphs discovered in previous builds
    annotated with file hashes, and the content-addressable cache.
\end{itemize}

### 2018-03-09 paper/compute.tex
Add draft of the background section
Restructure the paper

\section{Compute}\label{sec-compute}

\begin{figure}
\begin{minted}{haskell}
type IdentityCompute    k v = forall f.                  (k -> f v) -> k -> f v
type FunctorialCompute  k v = forall f. Functor     f => (k -> f v) -> k -> f v
type ApplicativeCompute k v = forall f. Applicative f => (k -> f v) -> k -> f v
type AlternativeCompute k v = forall f. Alternative f => (k -> f v) -> k -> f v
type MonadicCompute     k v = forall m. Monad       m => (k -> m v) -> k -> m v
type MonadPlusedCompute k v = forall m. MonadPlus   m => (k -> m v) -> k -> m v
\end{minted}
\caption{Compute abstractions}\label{fig-compute}
\end{figure}

Consider abstractions in Fig.~\ref{fig-compute}.

>>>>

\section{Build tasks, abstractly}\label{sec-compute}

This section presents purely functional abstractions for build systems that
allow us to express all the intricacies of build systems discussed in the
previous section~\S\ref{sec-background} and design complex build systems from
simple primitives. Specifically, we present the \emph{compute} and \emph{build}
abstractions:

\begin{minted}[xleftmargin=10pt]{haskell}
type Compute c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
type Build c i k v = Compute c k v -> k -> Maybe i -> Map k v -> (i, Map k v)
\end{minted}

\noindent
Compute represents build rules (in build systems) or formulas (in spreadsheets),
and is completely isolated from the world of compilers, file systems, memories,
caches, and all other complexities of real build systems. We explain the compute
type in~\S\ref{sec-general-compute}, and then discuss several noteworthy special cases:
functorial (\S\ref{sec-compute-functor}), applicative
(\S\ref{sec-compute-applicative}) and monadic (\S\ref{sec-compute-monad}) types
of compute.

Build corresponds to the algorithm used for bringing a key-value store to a
consistent state by recomputing dirty values in an appropriate order. We explain
the build type in~\S\ref{sec-build}, and then provide implementations for several
build systems in~\S\ref{sec-implementations}.

\subsection{Compute}\label{sec-general-compute}

...
% \begin{figure}
% \begin{minted}{haskell}
% type Compute c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
% \end{minted}
% \caption{Compute abstractions}\label{fig-compute}
% \end{figure}
% Consider abstractions in Fig.~\ref{fig-compute}.

\subsection{Functorial compute}\label{sec-compute-functor}

The \emph{Collatz sequence} $C_i$ is defined as follows:

\[
C_{i} = {\begin{cases}~n&{\text{for }}i=0\\~f(C_{i-1})&{\text{otherwise}},\end{cases}}\hspace{12pt}\text{where}\hspace{12pt}f(k)={\begin{cases}~k/2&{\text{if }}k\text{ is even}\\~3k+1&{\text{otherwise}}\end{cases}}
\vspace{2mm}
\]
\noindent
and $n$ is a positive integer parameter. The famous \emph{Collatz conjecture}
states that the Collatz sequence eventually reaches 1 for all possible values of
$n$. For example, if $n=6$, we reach 1 in eight steps:
$(6, 3, 10, 5, 16, 8, 4, 2, 1, \dots)$, after which the sequence loops forever:
$(4, 2, 1, 4, 2, 1, \dots)$.

We can express the computation of values in the Collatz sequence as a functorial
compute:

\begin{minted}[xleftmargin=10pt]{haskell}
data Collatz = Collatz Int

collatz :: Compute Functor Collatz Int
collatz get (Collatz k) | k <= 0    = Nothing
                        | otherwise = Just $ f <$> get (Collatz (k - 1))
  where
    f n | even n    = n `div` 2
        | otherwise = 3 * n + 1
\end{minted}

...

\subsection{Applicative compute}\label{sec-compute-applicative}
The \emph{generalised Fibonacci sequence} $F_i$ is defined as follows:

\[
F_{i} = {\begin{cases}~n&{\text{for }}i=0\\~m&{\text{for }}i=1\\~F_{i-1}+F_{i-2}&{\text{otherwise}}\end{cases}}
\vspace{2mm}
\]
\noindent
where $n$ and $m$ are integer parameters. By setting $n=0$ and $m=1$ we obtain
the famous \emph{Fibonacci sequence}: $(0, 1, 1, 2, 3, 5, 8, 13, \dots$), and if
$n=2$ and $m=1$, the result is the \emph{Lucas sequence}:
$(2, 1, 3, 4, 7, 11, 18, 29, \dots)$.

We can express the computation of values in the generalised Fibonacci sequence
as an applicative compute:

\begin{minted}[xleftmargin=10pt]{haskell}
data Fibonacci = Fibonacci Int

fibonacci :: Compute Applicative Fibonacci Int
fibonacci get (Fibonacci k) | k <= 1    = Nothing
                            | otherwise = Just $ (+) <$> get (Fibonacci (k - 1))
                                                     <*> get (Fibonacci (k - 2))
\end{minted}

...

\subsection{Monadic compute}\label{sec-compute-monad}

The \emph{Ackermann function} $A(m, n)$ is defined as follows:

\[
A(m, n) = {\begin{cases}~n+1&{\text{for }}m=0\\~A(m-1, 1)&{\text{for }}n=0\\~A(m-1,A(m,n-1))&{\text{otherwise}}\end{cases}}
\vspace{2mm}
\]
\noindent
We can express the computation of the Ackermann function as a monadic compute:

\begin{minted}[xleftmargin=10pt]{haskell}
data Ackermann = Ackermann Int Int

ackermann :: Compute Monad Ackermann Int
ackermann get (Ackermann m n)
    | m < 0 || n < 0 = Nothing
    | m == 0    = Just $ return (n + 1)
    | n == 0    = Just $ get (Ackermann (m - 1) 1)
    | otherwise = Just $ do
        index <- get (Ackermann m (n - 1))
        get (Ackermann (m - 1) index)
\end{minted}

...

\subsection{Build}\label{sec-general-build}

...

### 2018-03-09 paper/implementations.tex
Restructure the paper

\clearpage
\section{Build systems, concretely}\label{sec-examples}

In the previous sections we've discussed the types of build system in text, but the way we came to our conclusions was by developing a framework to fit the builds in. In this section we explain the framework, some of the abstractions we introduced (which model the properties from \S\ref{todo}), and how we can make a meaningful implementation of the build systems discussed in \S\ref{sec-background}.

@@

\subsection{\Make}

@@

\subsection{Spreadsheet}

@@

\subsection{Shake}

@@

\subsection{Bazel}

@@

The state matches that described in \S\ref{todo} - we have a map indexed by a target key, and the hashes of the dependencies of that key. Given that, we point at a resultant hash. Separately we have a map giving the contents for a given hash. For simplicity we assume that all hashes have corresponding entries in the content index.

@@

\subsection{Cloud Shake}

>>>>

\clearpage
\section{Build systems, concretely}\label{sec-implementations}

In the previous sections we've discussed the types of build system in text, but the way we came to our conclusions was by developing a framework to fit the builds in. In this section we explain the framework, some of the abstractions we introduced (which model the properties from \S\ref{sec-background}), and how we can make a meaningful implementation of the build systems discussed in \S\ref{sec-background}.

@@

\subsection{\Make}\label{sec-implementation-make}

@@

\subsection{\Excel}\label{sec-implementation-excel}

@@

\subsection{\Shake}\label{sec-implementation-shake}

@@

\subsection{\Bazel}\label{sec-implementation-bazel}

@@

The state matches that described in \S\ref{sec-build} - we have a map indexed by a target key, and the hashes of the dependencies of that key. Given that, we point at a resultant hash. Separately we have a map giving the contents for a given hash. For simplicity we assume that all hashes have corresponding entries in the content index.

@@

\subsection{Cloud \Shake}\label{sec-implementation-cloud-shake}
